---
title: "Analyze HMDA Mortgage Data with Snowflake Cortex AI and Positron Assistant"
format: html
editor_options:
  chunk_output_type: console
execute:
  echo: false
  warning: false
  error: false
---

The Snowflake guide this document accompanies:

- Uses the Home Mortgage Disclosure Act (HMDA) dataset from Snowflake's free public data
- Leverages Cortex AI with Databot and Positron Assistant for exploratory data analysis in Posit Workbench
- Builds an interactive Shiny dashboard with traditional UI controls
- Deploys to Posit Connect for easy and secure sharing and collaboration

## Install packages

```{r}
install.packages("renv")
```

```{r}
renv::activate()
renv::restore()
```

## Connect to data

```{r}
library(DBI)
library(odbc)
library(dplyr)
library(dbplyr)
library(connectcreds)

get_connection <- function() {
  warehouse <- "DEFAULT_WH"
  database <- "SNOWFLAKE_PUBLIC_DATA_FREE"
  schema <- "PUBLIC_DATA_FREE"
  account <- Sys.getenv("SNOWFLAKE_ACCOUNT")

  con <- DBI::dbConnect(
    odbc::snowflake(),
    account = account,
    warehouse = warehouse,
    database = database,
    schema = schema
  )

con <- get_connection()
mortgage_data <- tbl(con, "HOME_MORTGAGE_DISCLOSURE_ATTRIBUTES")
message("Successfully established secure connection to Snowflake!")
}
```

## Exploratory data analysis with Databot

Let's step out of our Quarto document and open Databot to explore our mortgage data. Come back when you're ready to build the dashboard.

## Build Interactive Dashboard

### Build a Shiny app

```{r}
# Create app.R file with Shiny application code
app_code <- '
library(shiny)
library(DBI)
library(odbc)
library(dplyr)
library(dbplyr)
library(connectcreds)
library(ggplot2)
library(scales)

get_connection <- function() {
  warehouse <- "DEFAULT_WH"
  database <- "SNOWFLAKE_PUBLIC_DATA_FREE"
  schema <- "PUBLIC_DATA_FREE"
  account <- Sys.getenv("SNOWFLAKE_ACCOUNT")

  con <- DBI::dbConnect(
    odbc::snowflake(),
    account = account,
    warehouse = warehouse,
    database = database,
    schema = schema
  )

  return(con)
}

# Define UI
ui <- fluidPage(
  titlePanel("HMDA Mortgage Data Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("loan_type", "Select Loan Type:",
                  choices = c("All", "Conventional", "FHA", "VA", "FSA/RHS")),
      selectInput("state", "Select State:",
                  choices = c("All", "California", "Texas", "Florida", "New York", "Pennsylvania")),
      sliderInput("loan_amount", "Loan Amount Range:",
                  min = 0, max = 1000000, value = c(0, 500000),
                  step = 50000)
    ),
    mainPanel(
      h4("Loan Distribution:"),
      plotOutput("loan_plot"),
      h4("Key Metrics:"),
      verbatimTextOutput("key_metrics"),
      h4("Year Trends:"),
      plotOutput("year_plot")
    )
  )
)

# Define server
server <- function(input, output, session) {
  # Initialize connection
  con <- get_connection()
  mortgage_data <- tbl(con, "HOME_MORTGAGE_DISCLOSURE_ATTRIBUTES")

  # Close connection when session ends
  onStop(function() {
    dbDisconnect(con)
  })

  # Reactive filtered data
  filtered_data <- reactive({
    # Extract slider values into local variables before using in filter
    min_amount <- input$loan_amount[1]
    max_amount <- input$loan_amount[2]

    data <- mortgage_data %>%
      filter(LOAN_AMOUNT >= min_amount,
             LOAN_AMOUNT <= max_amount)

    if (input$loan_type != "All") {
      data <- data %>% filter(LOAN_TYPE == input$loan_type)
    }

    if (input$state != "All") {
      data <- data %>% filter(STATE_NAME == input$state)
    }

    # Limit to 10,000 rows for performance
    result <- data %>%
      head(10000) %>%
      collect()

    result
  })

  output$loan_plot <- renderPlot({
    data <- filtered_data()

    if (nrow(data) == 0) {
      plot.new()
      text(0.5, 0.5, "No data available for the selected filters", cex = 1.5)
    } else {
      ggplot(data, aes(x = LOAN_AMOUNT)) +
        geom_histogram(bins = 30, fill = "steelblue", color = "white") +
        scale_x_continuous(labels = label_dollar(scale_cut = cut_short_scale())) +
        theme_minimal() +
        labs(title = paste("Distribution of Loan Amounts (", nrow(data), "loans)"),
             x = "Loan Amount",
             y = "Count")
    }
  })

  output$key_metrics <- renderPrint({
    data <- filtered_data()
    if (nrow(data) == 0) {
      cat("No data available for the selected filters")
    } else {
      total_loans <- nrow(data)
      avg_loan <- mean(data$LOAN_AMOUNT, na.rm = TRUE)
      total_volume <- sum(data$LOAN_AMOUNT, na.rm = TRUE)

      cat(sprintf("Total Loans: %s\\n", format(total_loans, big.mark = ",")))
      cat(sprintf("Average Loan Amount: %s\\n", dollar(avg_loan)))
      cat(sprintf("Total Loan Volume: %s\\n", dollar(total_volume)))
    }
  })

  output$year_plot <- renderPlot({
    data <- filtered_data()

    if (nrow(data) == 0) {
      plot.new()
      text(0.5, 0.5, "No data available for the selected filters", cex = 1.5)
    } else {
      year_summary <- data %>%
        group_by(YEAR) %>%
        summarise(count = n(), .groups = "drop")

      ggplot(year_summary, aes(x = factor(YEAR), y = count)) +
        geom_col(fill = "steelblue") +
        theme_minimal() +
        labs(title = "Loan Count by Year",
             x = "Year",
             y = "Number of Loans") +
        scale_y_continuous(labels = label_comma())
    }
  })
}

# Run the application
shinyApp(ui = ui, server = server)
'

# Write the app code to app.R file
writeLines(app_code, "app.R")

message("Shiny app created successfully!")
message("\nRun the app with: shiny::runApp(\"app.R\")")
```
